---
title: "Homework 4"
author: "Emma Sayre"
date: "May 13, 2026"
format: html
editor: visual
---

```{r}
#| label: load_libs
#| include: false
#| echo: false
#| message: false

library(tidyverse)
library(here)
library(janitor)

```

## Contract

**Input:**

n - total number of organisms of a particular species

N - total number of organisms of all species

**Output:**

D - Simpson's Diversity Index (range 0 to 1)

- Probability that two indiviudals randomly selected from a sample will belong to the same species

- D = 0 indiciates infinite diversity; D = 1 indicates no diversity

**Parameters:** None

**Formula:**

D = Σ \[n(n-1)\] / \[N(N-1)\]

## Implement Function 

```{r}
#| label: implement_function
#| include: true
#| echo: true
#| message: false

#' Simpson Biodiversity
#' 
#' @param n total number of orgs. of a species
#' @param N total number of orgs. of all species
#' @return D - Simpson's Diversity Index (0 to 1)
#' 
#' function definition 
biodiversity = function(n, N) {
  result = sum(n*(n-1)) / (N * (N-1))
  return(result)
}

```

## Test w/ Sample Data

```{r}
#| label: test
#| include: true
#| echo: true
#| message: true

#Generate sample data
trees = c("oak", "oak", "oak", "pine", "pine", "maple", "maple", "pine", "oak", "birch", "oak", "oak", "oak", "pine") 

species = as.factor(trees)
n = as.vector(table(species))
N = length (trees )

D = biodiversity(n,N)
```
