---
title: "Homework 3"
author: "Emma Sayre"
date: "May 7, 2026"
format: html
editor: visual
---

```{r}
#| label: load_libs
#| include: false
#| echo: false
#| message: false

#load libs
library(tidyverse)
library(here)
library(janitor)
library(dplyr)
library(ggplot2)
library(dplyr)
library(hrbrthemes)
```

## Part 1 - Build the Dataset

```{r}
#| label: build_dataset
#| include: true
#| echo: true
#| output: false
#| message: false

day <- 1:365

#create sample PM2.5 points
set.seed(125)
pm25 = runif(min=0, max=50, n=365)
pm25

#define seasons
season <- case_when(
  day %in% 1:90 ~ "winter",
  day %in% 91:180 ~ "spring",
  day %in% 181:270 ~ "summer",
  day %in% 271:365 ~ "fall"
) 

season <- as.factor(season)

#leave risk blank to populate later
risk = NA

#create air quality data frame
air_quality <- data.frame(
  day = day,
  pm25 = pm25,
  season = season,
  risk = risk)
air_quality 

summary(air_quality)
```

## Part 2 - Classify Risk with a For Loop

```{r}
#| label: for_loop
#| include: true
#| echo: true
#| message: false
#| output: false

#set variables
risk = air_quality$pm25
unhealthy = 35.4 #or above
good = 12 #below 12
num_unhealthy = 0
num_good = 0
longest_streak = 0
current_streak = 0

#Loop

for(i in 1:length(risk)) {
  #Count unhealthy and good days 
  num_unhealthy = num_unhealthy + ifelse(risk[i] >= unhealthy, 1, 0)
  num_good = num_good + ifelse(risk[i] < good, 1, 0) 
  
  #Populate risk column 
  if(risk[i] >= unhealthy) {
    air_quality$risk[i] = "unhealthy"
  } else if(risk[i] < good) {
    air_quality$risk[i] = "good"
  } else {
    air_quality$risk[i] = "moderate"
  }
  
  #Track longest streak
  if(risk[i] >= unhealthy) {
    current_streak = current_streak + 1
    longest_streak = max(longest_streak, current_streak)
  } else {
    current_streak = 0
    }
}

max_unhealthy_streak <- longest_streak
air_quality
```

## Part 3 - Seasonal Summary with a For Loop

```{r}
#| label: seasonal_summary 
#| include: true
#| echo: true
#| message: false

#Find the number of days in each season that fall into unhealthy risk category

unhealthy_winter = 0
unhealthy_spring = 0
unhealthy_summer = 0
unhealthy_fall = 0

for(i in 1:nrow(air_quality)) {
  if(air_quality$risk[i] == "unhealthy" & air_quality$season[i] == "winter") {
    unhealthy_winter = unhealthy_winter + 1
  } else if(air_quality$risk[i] == "unhealthy" & air_quality$season[i] == "spring") {
    unhealthy_spring = unhealthy_spring + 1
  } else if(air_quality$risk[i] == "unhealthy" & air_quality$season[i] == "summer") {
    unhealthy_summer = unhealthy_summer + 1
  } else if(air_quality$risk[i] == "unhealthy" & air_quality$season[i] == "fall") {
    unhealthy_fall = unhealthy_fall + 1
  } }

unhealthy_winter 
unhealthy_spring 
unhealthy_summer 
unhealthy_fall 

```

## Part 4 - Visualize and Report

###### Figure 1:

```{r}
#| label: time_series
#| include: true
#| echo: false
#| message: true
#| warnings: false

# Add months to air_quality dataframe
air_quality$date <- as.Date(air_quality$day -1, origin= "2025-01-01")

#Time Series Plot

pm25_plot <- ggplot(air_quality, 
                    aes(x=date, y=pm25, color = season)) +
  geom_line() +
  scale_color_manual(
    name = "Season",
    values = c("winter"= "skyblue3",
                                "spring" = "forestgreen",
                                "summer" = "salmon2",
                                "fall" = "goldenrod2"),
    breaks = c("winter", "spring", "summer", "fall"),
    labels = c("Winter", "Spring", "Summer", "Fall")) +
  xlab("") +
  ylab("PM2.5 Concentration (µg/m³)") +
  scale_x_date(date_labels = "%b") +
  ggtitle("Los Angeles PM2.5 Concentrations in 2025", 
          subtitle= sprintf("Longest Unhealthy Air Quality Streak: %d days", max_unhealthy_streak)) +
  theme_bw()
pm25_plot

```

###### Figure 2:

```{r}
#| label: bar_chart
#| include: true
#| echo: false
#| message: false
#| warning: false

bar_chart <- ggplot(air_quality, aes(x=risk, fill = risk)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  scale_fill_manual(
    name = "Risk Category",
    values = c("good"= "forestgreen",
               "moderate" ="goldenrod2",
               "unhealthy" = "indianred3"),
    breaks = c("good", "moderate", "unhealthy"),
    labels = c("Good", "Moderate", "Unhealthy")) +
  scale_x_discrete(limits = c("good", "moderate", "unhealthy"),
                   labels = c("Good", "Moderate", "Unhealthy")) +
  xlab("") +
  ylab("Number of Days") +
  ggtitle("Total Number of Days in Each Risk Category", 
          subtitle = sprintf("Longest Unhealthy Air Quality Streak: %d days", max_unhealthy_streak)) +
  theme_bw()
bar_chart

```

#### Summary of Findings

According to Figure 1, PM2.5 concentrations in Los Angeles fluctuated throughout all of 2025 despite the season, ranging from 0.01-49.83 µg/m³. The average concentration of PM2.5 was 26.08 µg/m³, falling into the moderate air quality range. Figure 2 displays this information further by showing that 172 days in 2025 fell into the moderate range, 119 were unhealthy, and 74 were good. Overall, Los Angeles is seeing few good air quality days in comparison to moderate and unhealthy days, indicating the potential for serious public health consequences. PM2.5 in the air can enter people's lungs and bloodstreams, causing strokes, heart attacks, asthma, respiratory disease, lung cancer, and even death. Thus, environmental managers should work to identify major pollution sources and coordinate with politicians to implement policies such as emission caps.

```{r}
#| label: stats
#| include: false
#| echo: false
#| message: false
#| warning: false

min(air_quality$pm25)
max(air_quality$pm25)
mean(air_quality$pm25)
```
